\documentclass[11pt,a4paper]{article}

% ── Geometry & layout ──────────────────────────────────────────────
\usepackage[margin=2.5cm]{geometry}
\usepackage{parskip}            % paragraph spacing instead of indent
\usepackage{setspace}
\onehalfspacing

% ── Math & symbols ─────────────────────────────────────────────────
\usepackage{amsmath,amssymb}

% ── Graphics & tables ──────────────────────────────────────────────
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{float}
\usepackage[font=small,labelfont=bf]{caption}

% ── Colors & code ──────────────────────────────────────────────────
\usepackage{multirow}
\usepackage[dvipsnames]{xcolor}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!8},
  frame=single,
  rulecolor=\color{gray!30},
  breaklines=true,
  numbers=left,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{MidnightBlue},
  commentstyle=\color{ForestGreen},
  stringstyle=\color{BrickRed},
  language=Python,
  showstringspaces=false,
}

% ── Headers & footers ─────────────────────────────────────────────
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textit{Solar PV Forecasting -- Technical Note}}
\fancyhead[R]{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ── Links ──────────────────────────────────────────────────────────
\usepackage[
  colorlinks=true,
  linkcolor=MidnightBlue,
  citecolor=MidnightBlue,
  urlcolor=MidnightBlue
]{hyperref}

% ── Lists ──────────────────────────────────────────────────────────
\usepackage{enumitem}
\setlist{nosep,leftmargin=1.5em}

% ── Bibliography ───────────────────────────────────────────────────
\usepackage[round,sort&compress]{natbib}

% ══════════════════════════════════════════════════════════════════
\begin{document}

% ── Title ──────────────────────────────────────────────────────────
\begin{center}
  {\LARGE\bfseries Quarter-Hourly National Solar PV Forecasting\\
  for Germany}\\[0.8em]
  {\large Technical Note}\\[0.6em]
  {\normalsize Xiyun Fu}\\[0.3em]
  {\small \today}
\end{center}

\vspace{0.5em}
\hrule
\vspace{1.5em}

% ── Abstract ───────────────────────────────────────────────────────
\begin{abstract}
\noindent
This report describes an end-to-end pipeline for predicting quarter-hourly
national solar photovoltaic (PV) generation in Germany. The pipeline combines
publicly available weather reanalysis data (ERA5 via Open-Meteo), a
capacity-weighted spatial distribution of installed PV across German federal
states, and realised generation data from the SMARD platform. A synthetic
national PV proxy is constructed by aggregating surface solar radiation across
regions weighted by installed capacity. Two forecasting models are evaluated:
a linear baseline regressing on the synthetic proxy, and a LightGBM model
incorporating calendar, solar position, and meteorological features. Both are
validated on a rolling-origin time-aware split using normalised MAE and a
skill score relative to the baseline. The pipeline targets full-year backfill
within 45 minutes on a reference VM (8~vCPU, 32~GB RAM) and outputs
predictions in Parquet format via a single command.
\end{abstract}

% ── Table of contents (optional, remove if too short) ──────────────
\tableofcontents
\vspace{1em}

% ══════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}

Accurate forecasting of solar PV generation is critical for grid balancing,
intraday trading, and reserve dimensioning. Germany, with over 80~GW of
installed PV capacity as of 2024, presents a challenging national-scale
aggregation problem: generation depends on spatially heterogeneous irradiance
conditions across diverse latitudes and orientations.

This work addresses the task of predicting quarter-hourly (15-minute)
national solar PV production for Germany by combining three data layers:
\begin{enumerate}
  \item A \textbf{weather/irradiance layer} providing surface solar radiation
        and meteorological covariates;
  \item A \textbf{PV spatial distribution layer} capturing the geographic
        allocation of installed capacity;
  \item A \textbf{realised generation series} serving as the prediction target.
\end{enumerate}

The pipeline is designed to run reproducibly on a modest compute budget
(8~vCPU, 32~GB RAM, no GPU) with a target wall-clock time of approximately
45~minutes for a full-year backfill. All code, data pipelines, and model
artefacts are provided in a public repository with a single-command entry
point (\texttt{make full-run}).

Section~\ref{sec:data} describes data sourcing and engineering decisions.
Section~\ref{sec:proxy} details the synthetic proxy construction.
Section~\ref{sec:model} presents the modelling and validation framework.
Section~\ref{sec:results} reports results. Section~\ref{sec:discussion}
discusses limitations and next steps.

% ══════════════════════════════════════════════════════════════════
\section{Data Sourcing \& Engineering}
\label{sec:data}

\subsection{Data sources considered}
\label{sec:data:considered}

Table~\ref{tab:datasources} summarises the datasets considered, those
selected, and the rationale for each decision.

\begin{table}[H]
\centering
\caption{Data sources considered and selection rationale.}
\label{tab:datasources}
\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Layer} & \textbf{Source} & \textbf{Used?} & \textbf{Rationale} \\
\midrule
\multirow{3}{*}{Weather}
  & ERA5 (CDS direct)        & No  & Requires CDS credentials; slow download \\
  & Open-Meteo Historical    & Yes & Free; ERA5 backend; hourly; no API key \\
  & DWD ICON / COSMO         & No  & Forecast data, not reanalysis \\
\midrule
\multirow{3}{*}{PV spatial}
  & MaStR (BNetzA)           & No  & Plant-level; complex parsing; overkill \\
  & OPSD capacity data       & Yes & Aggregated by state; well-documented \\
  & OSM-derived              & No  & Incomplete coverage \\
\midrule
\multirow{3}{*}{Target}
  & ENTSO-E (entsoe-py)      & No  & Requires API key (3-day wait) \\
  & SMARD (BNetzA)           & Yes & Direct CSV; 15-min; no registration \\
  & OPSD time series         & No  & May lag behind most recent year \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Weather / irradiance layer}
\label{sec:data:weather}

We retrieve hourly ERA5-derived data from the Open-Meteo Historical Weather
API for a set of representative coordinates across Germany's 16 federal
states. For each location we request:
\begin{itemize}
  \item \textbf{shortwave\_radiation} (GHI, W/m\textsuperscript{2}) --
        surface incoming solar radiation;
  \item \textbf{temperature\_2m} (\textdegree C) -- affects PV cell efficiency;
  \item \textbf{windspeed\_10m} (m/s) -- contributes to convective cooling.
\end{itemize}

Open-Meteo serves ERA5 reanalysis data at 0.25\textdegree\ spatial resolution
($\sim$25~km). We select one coordinate per Bundesland, chosen to
approximate the centroid of the state's PV-dense area, avoiding coastal
or alpine extremes.

\textbf{Temporal resolution.} ERA5 provides hourly data. We interpolate to
15-minute resolution using cubic spline interpolation for GHI (preserving
non-negativity via clipping) and linear interpolation for temperature and
wind speed.

% TODO: insert figure showing representative coordinates on map

\subsection{PV spatial distribution layer}
\label{sec:data:pv}

We use aggregated installed PV capacity per German federal state from the
Open Power System Data (OPSD) platform. This provides a well-documented
proxy for the spatial distribution of installed capacity without requiring
plant-level resolution.

The capacity weights are normalised:
\begin{equation}
  w_i = \frac{C_i}{\sum_{j \in \mathcal{S}} C_j}
  \label{eq:weights}
\end{equation}
where $C_i$ is the installed PV capacity in state $i$ and $\mathcal{S}$ is
the set of all states.

% TODO: insert table with per-state capacity and weights

\subsection{Target series}
\label{sec:data:target}

Realised German solar generation is sourced from SMARD
(\url{https://smard.de}), operated by the Bundesnetzagentur. The data is
available at 15-minute resolution in MW and covers generation from all
four German TSO zones (50Hertz, Amprion, TenneT, TransnetBW) aggregated
to the national level.

We download data for January--December 2025 (or the latest available
month). Timestamps are in CET/CEST; we convert to UTC for internal
consistency.

\subsection{Harmonisation}
\label{sec:data:harmonise}

All three layers are aligned to a common schema:
\begin{itemize}
  \item \textbf{Timestamps}: 15-minute intervals, UTC, timezone-naive
        \texttt{datetime64[ns]};
  \item \textbf{Units}: GHI in W/m\textsuperscript{2}, generation in MW,
        capacity in MW\textsubscript{p};
  \item \textbf{Missing values}: gaps $<$4 hours are forward-filled;
        longer gaps are flagged and excluded from training;
  \item \textbf{Night masking}: periods where solar zenith angle $>$90\textdegree\
        are set to zero.
\end{itemize}

Intermediate outputs are persisted as Parquet files partitioned by
year/month with \texttt{float32} precision where safe. Processing uses
\texttt{polars} with lazy evaluation to keep peak memory well below 32~GB.

% ══════════════════════════════════════════════════════════════════
\section{Synthetic Proxy Construction}
\label{sec:proxy}

The synthetic national PV proxy aggregates regional irradiance into a
single national time series:
\begin{equation}
  P_t^{\text{syn}} = \eta \sum_{i \in \mathcal{S}} \text{GHI}_{i,t} \cdot w_i
  \label{eq:proxy}
\end{equation}
where $\text{GHI}_{i,t}$ is the surface solar radiation at the
representative coordinate for state $i$ at time $t$, $w_i$ is the
normalised capacity weight from Eq.~\eqref{eq:weights}, and $\eta$ is
a scalar efficiency factor fitted by ordinary least squares on the
training set to minimise the residual against realised generation.

This formulation is intentionally simple. It assumes:
\begin{enumerate}
  \item PV generation scales linearly with GHI (ignoring temperature and
        angle-of-incidence effects);
  \item One coordinate per state is sufficient to capture regional
        irradiance variation;
  \item Capacity is static over the forecast period.
\end{enumerate}

These simplifications are justified by the scope of this exercise.
Section~\ref{sec:discussion} discusses how each could be relaxed.

% TODO: insert scatter plot of P_syn vs actual generation

% ══════════════════════════════════════════════════════════════════
\section{Modelling \& Validation}
\label{sec:model}

\subsection{Feature engineering}
\label{sec:model:features}

Beyond the raw synthetic proxy, we engineer the following features:
\begin{itemize}
  \item \textbf{Calendar}: hour-of-day, day-of-year (cyclically encoded
        as sin/cos), weekday indicator;
  \item \textbf{Solar position}: solar zenith angle, solar azimuth,
        clear-sky GHI (computed via \texttt{pvlib});
  \item \textbf{Clear-sky index}: $k_t = \text{GHI}_{\text{actual}} /
        \text{GHI}_{\text{clearsky}}$, clipped to $[0, 1.5]$;
  \item \textbf{Meteorological}: temperature, wind speed (national
        weighted averages);
  \item \textbf{Lags}: 1-day and 7-day lagged actual generation
        (only for non-operational forecasting; discussed as leakage
        risk in Section~\ref{sec:model:validation}).
\end{itemize}

\subsection{Baseline model}
\label{sec:model:baseline}

The baseline is a ridge regression:
\begin{equation}
  \hat{P}_t = \beta_0 + \beta_1 P_t^{\text{syn}} + \beta_2 k_t + \varepsilon_t
\end{equation}
fitted on the training fold. This provides a transparent, explainable
reference point.

\subsection{Candidate model}
\label{sec:model:candidate}

The candidate model is a \textbf{LightGBM} gradient-boosted tree
regressor using all engineered features. Key hyperparameters:
\begin{itemize}
  \item \texttt{num\_leaves}: 63
  \item \texttt{learning\_rate}: 0.05
  \item \texttt{n\_estimators}: 500 (with early stopping, patience 50)
  \item \texttt{min\_child\_samples}: 50
  \item \texttt{subsample}: 0.8
\end{itemize}

LightGBM is chosen for its computational efficiency (compatible with the
no-GPU constraint), native handling of missing values, and strong
empirical performance on tabular regression tasks.

\subsection{Validation setup}
\label{sec:model:validation}

We use a \textbf{rolling-origin} (expanding window) cross-validation
with monthly test blocks:
\begin{itemize}
  \item Initial training window: January--June 2025;
  \item Test blocks: each subsequent month evaluated independently;
  \item No overlap between training and test.
\end{itemize}

\textbf{Primary metric}: normalised MAE (nMAE), defined as:
\begin{equation}
  \text{nMAE} = \frac{\text{MAE}}{\overline{P}_{\text{actual}}}
  = \frac{\frac{1}{N}\sum_t |P_t - \hat{P}_t|}{\frac{1}{N}\sum_t P_t}
\end{equation}
where the denominator is the mean actual generation over the test period
(excluding nighttime zeros).

\textbf{Secondary metric}: skill score relative to the baseline:
\begin{equation}
  \text{SS} = 1 - \frac{\text{MAE}_{\text{candidate}}}{\text{MAE}_{\text{baseline}}}
\end{equation}

\textbf{Leakage controls}:
\begin{itemize}
  \item No future weather data in features (all features are concurrent
        or lagged);
  \item Lag features use only realised values from $\geq$24 hours prior;
  \item Train/test split is strictly temporal (no shuffling).
\end{itemize}

% ══════════════════════════════════════════════════════════════════
\section{Results}
\label{sec:results}

% TODO: fill in after running experiments

\begin{table}[H]
\centering
\caption{Out-of-sample performance (rolling-origin validation).}
\label{tab:results}
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Model}   & \textbf{MAE (MW)} & \textbf{nMAE (\%)} & \textbf{Skill Score} \\
\midrule
Baseline (ridge)  & --   & --  & --    \\
LightGBM          & --   & --  & --    \\
\bottomrule
\end{tabular}
\end{table}

% TODO: insert time-series plot of predictions vs actuals
% TODO: insert residual analysis (diurnal pattern, seasonal)
% TODO: insert feature importance plot for LightGBM

% ══════════════════════════════════════════════════════════════════
\section{Discussion}
\label{sec:discussion}

\subsection{Limitations}

\begin{enumerate}
  \item \textbf{Single coordinate per state}: sub-state irradiance
        variability is not captured. Larger states (e.g.\ Bavaria,
        North Rhine-Westphalia) would benefit from multiple sample
        points.
  \item \textbf{Hourly weather interpolated to 15-min}: the weather
        layer's native resolution is hourly, introducing smoothing at
        the quarter-hourly target. Satellite-derived irradiance
        (e.g.\ Open-Meteo Satellite API at 15-min native) would
        improve fidelity.
  \item \textbf{Static capacity weights}: installed PV capacity grows
        year-on-year. Our weights are a snapshot and do not reflect
        intra-year commissioning.
  \item \textbf{No temperature correction}: cell temperature effects on
        efficiency are ignored in the proxy. A simple temperature
        coefficient ($\sim -0.4$\%/\textdegree C above 25\textdegree C)
        could improve accuracy.
  \item \textbf{No curtailment modelling}: periods of grid curtailment
        reduce actual generation below physical potential.
\end{enumerate}

\subsection{Next steps}

\begin{itemize}
  \item Replace single-point weather with gridded ERA5 or satellite data
        and use \texttt{atlite} for physically-informed PV conversion;
  \item Incorporate \texttt{pvlib} temperature models for cell-level
        efficiency;
  \item Add probabilistic forecasting (quantile regression via LightGBM)
        for uncertainty estimation;
  \item Explore temporal models (LSTM, TiDE) if longer historical data
        is available;
  \item Integrate real-time data feeds for operational forecasting.
\end{itemize}

% ══════════════════════════════════════════════════════════════════
\section{Reproducibility}
\label{sec:reproducibility}

The full pipeline is available at:
\begin{center}
  \url{https://github.com/[username]/solar-pv-forecast}
\end{center}

\textbf{Environment}: Python $\geq$3.11, managed via \texttt{uv}. All
dependencies are locked in \texttt{pyproject.toml}.

\textbf{Single-command execution}:
\begin{lstlisting}[language=bash]
make full-run  # produces predictions.parquet + runtime log
\end{lstlisting}

\textbf{Runtime}: approximately [XX] minutes on the reference VM
(8~vCPU, 32~GB RAM). Peak memory: [XX]~GB.

% ══════════════════════════════════════════════════════════════════
\section{Data Documentation}
\label{sec:datadoc}

\begin{table}[H]
\centering
\caption{Variable definitions and units.}
\label{tab:variables}
\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Variable} & \textbf{Unit} & \textbf{Resolution} & \textbf{Source} \\
\midrule
\texttt{ghi\_wm2}          & W/m\textsuperscript{2} & 15-min (interp.)  & Open-Meteo \\
\texttt{temperature\_2m}   & \textdegree C          & 15-min (interp.)  & Open-Meteo \\
\texttt{windspeed\_10m}    & m/s                    & 15-min (interp.)  & Open-Meteo \\
\texttt{actual\_solar\_mw} & MW                     & 15-min (native)   & SMARD \\
\texttt{proxy\_solar\_mw}  & MW                     & 15-min (computed) & Derived \\
\texttt{solar\_zenith}     & degrees                & 15-min (computed) & pvlib \\
\texttt{clearsky\_ghi}     & W/m\textsuperscript{2} & 15-min (computed) & pvlib \\
\texttt{clearsky\_index}   & dimensionless          & 15-min (computed) & Derived \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Coordinate reference system}: WGS84 (EPSG:4326) for all
geographic coordinates.

\textbf{Resampling}: hourly $\rightarrow$ 15-min via cubic spline (GHI)
and linear interpolation (temperature, wind). Nighttime GHI clamped to
zero.

\textbf{Storage format}: Apache Parquet, partitioned by year/month,
\texttt{float32} for weather variables, \texttt{float64} for generation.

% ══════════════════════════════════════════════════════════════════

% Uncomment when references are ready:
% \bibliographystyle{abbrvnat}
% \bibliography{references}

\end{document}
